---
title: "Untitled"
author: "MT"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}


##here is what you would run in terminal to setup an environment like the rstudio environment

#conda create --name RStudio python=3.8
#conda activate RStudio
#conda install pandas matplotlib seaborn scipy statsmodels 
#conda env create -f env_config.yml

#install.packages("reticulate")


####### Is this here


# or for manual setup 

# # Set Python path
#use_python(config$python_path)
# 
# # Install Python packages (optional)
# py_install(config$packages)
# 
# 
# reticulate::py_install("pandas")
# reticulate::py_install("matplotlib")
# reticulate::py_install("seaborn")
# reticulate::py_install("scipy")
# reticulate::py_install("statsmodels")

library(reticulate)
library(pandoc)



```

### Overview 

Data Loading and Preprocessing: Loaded the data from the Excel file, checked for missing values.

Basic Summary Statistics: Calculated the mean, standard deviation, and standard error for each genotype and incubation time, grouped by date.

Control Data Extraction: Separated the control data ('bNY30a_parent') and calculated its average, standard deviation, and standard error. This control data is essential for normalizing the other data points.

Error Propagation: Calculated the normalized average and proceeded with error propagation to get the propagated standard deviation and standard error for each genotype. This was done considering the control data.

Covariance and Correlated Errors: Introduced the concept of covariance to account for the relationship between the variable and control when they are not independent. This ensures a more accurate estimate of errors.

```{r}

date()
Sys.time()
getwd()
list.files()
.libPaths()
.Library
sessionInfo()
search()
searchpaths()
R.version

```

```{r, include = FALSE}

use_condaenv("RStudio")


```






```{python}
# Importing required libraries
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns
import scipy.stats as f_oneway
import matplotlib as mpl
import statsmodels.api as sm



```

```{python}
# Load the dataset from 'data_complete.xlsx'
file_path = 'Data/data_master.xlsx'
data_master = pd.read_excel(file_path)




```

```{python}
# Calculate summary statistics
summary_stats = data_master.groupby(['date', 'genotype']).agg(
    avg_titer=('titer', np.mean),
    stdev_titer=('titer', np.std),
    n=('titer', 'count')
).reset_index()
summary_stats['se_titer'] = summary_stats['stdev_titer'] / np.sqrt(summary_stats['n'])

# Extract control data
control_data = summary_stats[summary_stats['genotype'] == 'No Cell Control'].copy()
control_data.rename(columns={'avg_titer': 'control_avg_titer', 'stdev_titer': 'control_stdev_titer', 'se_titer': 'control_se_titer'}, inplace=True)
control_data.drop(columns=['genotype', 'n'], inplace=True)

# Merge control data with original summary statistics data based on 'date'
avg_data_with_controls = pd.merge(summary_stats, control_data, on='date')

# Calculate normalized average
avg_data_with_controls['normalized_avg'] = avg_data_with_controls['avg_titer'] / avg_data_with_controls['control_avg_titer']

# Error propagation for standard deviation
avg_data_with_controls['normalized_stdev_propagation'] = avg_data_with_controls['normalized_avg'] * np.sqrt(
    (avg_data_with_controls['stdev_titer'] / avg_data_with_controls['avg_titer']) ** 2 +
    (avg_data_with_controls['control_stdev_titer'] / avg_data_with_controls['control_avg_titer']) ** 2
)

# Standard error based on propagated standard deviation
avg_data_with_controls['normalized_se_propagation'] = avg_data_with_controls['normalized_stdev_propagation'] / np.sqrt(avg_data_with_controls['n'])



```



```{python}

# Importing required libraries for statistical analysis
from statsmodels.formula.api import ols
from statsmodels.stats.anova import anova_lm
from statsmodels.regression.mixed_linear_model import MixedLM

# Performing one-way ANOVA
anova_model = ols('normalized_avg ~ C(genotype)', data=avg_data_with_controls).fit()
anova_table = anova_lm(anova_model, typ=2)

# Display the ANOVA table
anova_table

```

```{python}

try:
    mixedlm_model = MixedLM.from_formula('normalized_avg ~ 1', groups='date', re_formula='1', data=avg_data_with_controls)
    mixedlm_result = mixedlm_model.fit()
    mixedlm_summary = mixedlm_result.summary()
except Exception as e:
    mixedlm_summary = str(e)

mixedlm_summary



```