---
title: "CpaF_mutant_EOP"
author: "MT"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Housekeeping

```{r}
date()
Sys.time()
getwd()
list.files()
.libPaths()
.Library
sessionInfo()
search()
searchpaths()
R.version


```

# Checking and installing packages

```{r, results = 'hide'}

packages <- c("ggplot2", "dplyr", "tidyr", "readxl", "ggpubr", "skimr", "DataExplorer", "tidyverse", "skimr", "svglite", "readxl","tidyxl", "ggforce", "ggpubr", "ggsci", "ggthemes", "ragg", "magrittr", "reprex", "patchwork", "writexl", 'magick', 'forcats')

# list of packages sometimes I use [   "searcher", "errorist",

install.packages(setdiff(packages, rownames(installed.packages())))

update.packages (packages)

for (i in 1:length(packages)) {

library(packages[i], character.only = TRUE)


}

```


# Importing the data from Master Excel Sheet

need to also rename from strain number to relevant genotype

```{r}

october28 <- read_excel("Data/Triplicate Plate Titer EOP for Caulobacter.xlsx", 
    range = "C3:E11", col_names = FALSE)

colnames(october28) <- c("strain", "plaques", "titer")

october28 <- mutate(october28, strain = as.character(strain), date = as.Date("2021-10-28"), incubation_time = as.factor(18))

##

november2 <- read_excel("Data/Triplicate Plate Titer EOP for Caulobacter.xlsx", 
    range = "C20:E43", col_names = FALSE)

colnames(november2) <- c("strain", "plaques", "titer")

november2 <- mutate(november2, strain = as.character(strain), date = as.Date("2021-11-2"), incubation_time = as.factor(18))

##

november8 <- read_excel("Data/Triplicate Plate Titer EOP for Caulobacter.xlsx", 
    range = "C50:E73", col_names = FALSE)

colnames(november8 ) <- c("strain", "plaques", "titer")

november8  <- mutate(november8 , strain = as.character(strain), date = as.Date("2021-11-8"), incubation_time = as.factor(18))

##

november10 <- read_excel("Data/Triplicate Plate Titer EOP for Caulobacter.xlsx", 
    range = "C81:E104", col_names = FALSE)

colnames(november10) <- c("strain", "plaques", "titer")

november10 <- mutate(november10, strain = as.character(strain), date = as.Date("2021-10-28"), incubation_time = as.factor(18))

## now for the 48 hours


november3 <- read_excel("Data/Triplicate Plate Titer EOP for Caulobacter.xlsx", 
    range = "J20:L46", col_names = FALSE)

colnames(november3) <- c("strain", "plaques", "titer")

november3 <- mutate(november3, strain = as.character(strain), date = as.Date("2021-11-03"), incubation_time = as.factor(48))

##

november9 <- read_excel("Data/Triplicate Plate Titer EOP for Caulobacter.xlsx", 
    range = "J50:L73", col_names = FALSE)

colnames(november9) <- c("strain", "plaques", "titer")

november9 <- mutate(november9, strain = as.character(strain), date = as.Date("2021-11-09"), incubation_time = as.factor(48))

##

november10_48 <- read_excel("Data/Triplicate Plate Titer EOP for Caulobacter.xlsx", 
    range = "J81:L106", col_names = FALSE)

colnames(november10_48) <- c("strain", "plaques", "titer")

november10_48 <- mutate(november10_48, strain = as.character(strain), date = as.Date("2021-11-10"), incubation_time = as.factor(48))

Rawdata <- rbind(november10, november10_48, november3, november8, november9, november2)

write_xlsx(Rawdata, "Data/RawData.xlsx")



```


```{r}

data1 <- Rawdata %>% 
  mutate(genotype = case_when(
    grepl("221", .$strain) ~ "bNY30a_parent",
    grepl("22222", .$strain) ~ "hfsA+",
    grepl("22223.", .$strain) ~ "wt",
    grepl("224", .$strain) ~ "delCpaF",
    grepl("225", .$strain) ~ "CpaF",
    grepl("226", .$strain) ~ "CpaF(ab)",
    grepl("228", .$strain) ~ "CpaF(I355C)",
    grepl("229", .$strain) ~ "CpaF(D310N)",
    grepl("230.", .$strain) ~ "CpaF(F244L/K245R)"
  )) 



unique(data1$genotype)
unique(data1$strain)

data2 <- data1 |> 
  mutate(genotype = as.factor(genotype)) |> 
  add_count(genotype, incubation_time) |>
  ungroup()

data_master <- data2

write_xlsx(data_master, "Data/data_master.xlsx")

```

## Traditional Statistics and transformations

```{r}
## firstly i will just calculate sd over all the different days, maybe later needs to be each individual day

## how to normalize a sample and include standard deviations or SEM
#Being B the control sample:
#A*=(A-mean(A))*std(B)/std(A)+mean(B)


eop_data <- data2 |> filter(incubation_time == 48) |> 
  group_by(genotype) |>
  mutate(
    titeraverage = mean(titer),
    stdev = sd(titer),
    sem = stdev/n,
    eop = titer/(3.64667e+11),
    eopaverage = mean(eop), 
    eopstdev = ((titer - titeraverage)*52.64266331)/(stdev+3.64667e+11)
    
  ) |>
  ungroup()


skim_eop_data <- eop_data |> group_by(genotype, incubation_time) |>
  skim() |>
  ungroup()

write_xlsx(skim_eop_data, "Data/skim_eop_data.xlsx")


```

### Exploring the data

```{r}

introduce(data_master)
plot_intro(data_master)
plot_missing(data_master)
plot_bar(data_master, by = "genotype")

pca_df <- na.omit(data_master)
pca_df <- drop_columns(pca_df, c(1))

plot_prcomp(pca_df)
plot_correlation(pca_df)

plot_bar(skim_data, by = "genotype")
```
## Doing arithmetic for EOP values and to check differences of 18 vs 48 hour incubations.

```{r}
skim_data <- data_master |> group_by(genotype, incubation_time) |>
  skim() |>
  ungroup()

write_xlsx(skim_data, "Data/skim_data.xlsx")



plot_data <- skim_data |>
  group_by(genotype, incubation_time) |>
  filter(skim_type == "numeric") |>
  mutate(sem = numeric.sd/(numeric.mean[skim_variable == "n"]), 
         counting_error = sqrt(numeric.mean[skim_variable == "plaques"])) |>
  ungroup() |>
  group_by(skim_variable, genotype) |>
  mutate(inceop = numeric.mean/numeric.mean[incubation_time == 48]) |>
  ungroup() |>
  group_by(skim_variable, incubation_time) |>
  mutate(eop = numeric.mean/numeric.mean[genotype == "CpaF"]) |>
  ungroup() |>
  select(skim_variable:incubation_time | numeric.mean:numeric.sd | sem:eop)

write_xlsx(plot_data, "Data/plot_data.xlsx")

```



## Plotting the EOP Data

```{r}

## how to normalize a sample and include standard deviations or SEM
#Being B the control sample:
#A*=(A-mean(A))*std(B)/std(A)+mean(B)

eop_barplot <- plot_data |> filter(skim_variable == "titer") |>
  ggplot(aes(x = genotype, y = eop, fill = incubation_time)) +
  geom_bar(stat = "identity", position = position_dodge())

eop_barplot

titer_boxplot <- ggplot(data_master, aes(x = genotype, y = titer, fill = incubation_time)) +
  geom_boxplot(outlier.size = 2, outlier.color = "black")+
  geom_point(color = "grey", size = 2, alpha = 0.5, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.7))

titer_boxplot

titer_barplot <- data3 |> 
  #filter(genotype %in% c(CpaF)) |>
  ggplot(aes(x = genotype, y = titeraverage)) +
  geom_bar(stat = "identity", position = position_dodge())+
  geom_errorbar(aes(ymin = titeraverage, ymax = titeraverage + stdev)) +
  scale_y_continuous(name = "Plaque forming units", trans = "log10", na.value = 0, n.breaks = 6) +
  coord_cartesian(ylim = c(1,1000000000000))
  
 

titer_barplot

titer_barplot_subset <- data3 |> 
  filter(genotype %in% c("CpaF", "CpaF(I355C)", "CpaF(D310N)", "CpaF(F244L/K245R)")) |>
  ggplot(aes(x = genotype, y = titeraverage)) +
  geom_bar(stat = "identity", position = position_dodge())+
  geom_errorbar(aes(ymin = titeraverage, ymax = titeraverage + stdev)) +
  scale_y_continuous(name = "Plaque forming units", trans = "log10", na.value = 0, n.breaks = 6) +
  coord_cartesian(ylim = c(1,1000000000000))

titer_barplot_subset

```
## Now lets look at the EOP values when only using the 48 hour incubations (less varibility based on barplots)


### Calculating the standard deviations based on using all replicates (population stdev)

```{r}
## how to normalize a sample and include standard deviations or SEM
#Being B the control sample:
#A*=(A-mean(A))*std(B)/std(A)+mean(B)

eop_datapoints <- data2 |> mutate(eop = titer/(3.64667e+11))

eop_data1 <- data2 |>filter(incubation_time == 48, genotype == c("CpaF", "CpaF(I355C)", "CpaF(D310N)", "CpaF(F244L/K245R)")) |>
  mutate(
    eop = titer/(3.64667e+11)
  )

eop_data <- data2 |> filter(incubation_time == 48, genotype == c("CpaF", "CpaF(I355C)", "CpaF(D310N)", "CpaF(F244L/K245R)")) |> 
  group_by(genotype) |>
  mutate(
    titeraverage = mean(titer),
    stdev = sd(titer),
    sem = stdev/n,
    eop = titer/(3.64667e+11),
    eopaverage = mean(eop), 
    eopstdev = ((titer - titeraverage)*52.64266331)/(stdev+3.64667e+11)
    
  ) |>
  ungroup()


skim_eop_data <- eop_data |> group_by(genotype, incubation_time) |>
  skim() |>
  ungroup()


eop_48_boxplot <- ggplot(eop_data, aes(x = genotype, y = eop)) +
  geom_boxplot(aes(fill = genotype), outlier.size = 2, outlier.color = "grey50")

eop_48_boxplot ## i am leaving off with this error

eop_boxplot <- eop_48_boxplot + theme_pubr_MT() + scale_color_npg() + theme(plot.margin = unit(c(1, 1, 1, 1), "mm"))

eop_boxplot_datapoints <- eop_boxplot + geom_point(data = )

```



```{r}



# Example dataset with a factor variable called "CpaF_variant"
df <- data.frame(
  value = rnorm(9),
  CpaF_variant = c("CpaF+", "CpaF+", "CpaF+", "delCpaF", "delCpaF", "delCpaF", "CpaF(I355C)", "CpaF(I355C)", "CpaF(I355C)")
)

# Subset the dataset to include only the "delCpaF" values
df_delCpaF <- filter(df, CpaF_variant == "delCpaF")

# Subset the dataset to include only the "CpaF+" and "CpaF(I355C)" values
df_CpaF <- filter(df, CpaF_variant %in% c("CpaF+", "CpaF(I355C)"))




```
